HIPPOCAMPAL SEGMENTATION WITH A VERY SMALL ATLAS LIBRARY: THE MAGeT BRAIN ALGORITHM

### 2012-11-01 paper.Rnw
Miriam and Julie comments incorporated. Changed all discussion of multi-atlas to "basic multi-atlas" to disambiguate. Squashed into four pages.

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. Unfortunately, these
approaches often rely on a large number of such manually segmentated atlases
that take significant time and expertise to produce. We present an algorithm
for the automatic segmentation of the hippocampus that minimizes the number of
atlases needed whilst still achieving similar accuracy to multi-atlas
approaches.  We perform repeated random subsampling validation on the IBSR
dataset to compare our approach to multi-atlas segmentation using the full IBSR
dataset, and to single-atlas (model-based) segmentation. Our results show that
with only 8 input atlas, MAGeT brain can achieve to within 2.0\% segmentation
accuracy of the multi-atlas approach using 17 input atlases (mean $\kappa =
0.775$ vs.  $\kappa = 0.791$).  These results compare favourably to others who
have used the IBSR data for validation. 
\end{abstract}

@@

Currently, the gold standard for neuroanatomical segmentation is manual
labelling by an expert human rater.  This is problematic for segmentation of
the hippocampus for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \cite{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true delineation of hippocampal anatomy in MR images is
disputed\cite{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol\cite{Jack2011}.  Compounding each of these problems is
the significant neuroanatomical variability in the hippocampus throughout the
course of aging, development, and neuropsychiatric disorders\cite{Mouiha2011}.

As well, it may be necessary to use several different hippocampal
definitions or, in fact, make research-specific modifications in order to test
hypotheses (for example, \cite{Poppenk2011} found that overall hippocampal
volume difference did not predict recollection memory performance but by
dividing the hippocampus into anterior and posterior regions, a predictive
volume difference was found).  Thus, whilst manual segmentation of the
hippocampus is an important technique, it may be a bottleneck for researchers
or clinicians who do not have access to the needed human expertise.

@@

The essential insight of generating a template library is not new.  Heckemann
\cite{Heckemann2006} compared generating a template library from a single atlas
to standard multi-atlas segmentation and found poor performance and so deemed
the approach as inviable.  The LEAP algorithm \cite{Wolz2010} proceeds by
iteratively segmenting the unlabelled image most similar to the atlas library
images and then incorporating the now-labelled image into the atlas library,
but requires 30 starting atlases.  The novelty of our method is to demonstrate
the possibility of producing comparable segmentation accuracy to these and
other multi-atlas-based methods whilst using significantly fewer manually
created atlases.

In previous work from our group \cite{Chakravarty2012}, we applied MAGeT brain
to the segmentation the striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. In the present work we extend our approach to the
hippocampus, a much more difficult structure to segment, and perform a thorough
validation over a range of atlas and template library sizes. Due to the small
number of atlases required, our method could easily accommodate different
hippocampal definitions. Our aim is not to improve on segmentation accuracy
beyond existing methods, but instead to provide a method that trades off manual
segmentation expertise for computational processing time whilst providing
sufficient accuracy for clinical and research applications.

@@

The segmentation approach we propose is best understood as an extension of
traditional multi-atlas segmentation.  In multi-atlas segmentation, an atlas
library and unlabelled MR images are given as input.  Every atlas image is
nonlinearly registered to each unlabelled image, and then each atlas' labels
are propagated via the resulting transformations.  These labels are then fused
to produce a single, definitive segmentation by some label fusion method (e.g.
voxel-wise majority vote).

Our extension adds a preliminary stage in which a template library is
constructed from input images, and used in place of an atlas library in the
standard multi-atlas-based method.  To create the template library, labels from
each atlas image are propagated to each template library image via the
transformation resulting from a non-linear registration between pair of images.
As a result, each template library image has a label from each atlas.
Traditional multi-atlas segmentation is then used to produce segmentations for
the entire set of unlabelled images (including those images used in the
template library).

@@

\subsubsection{Comparison Approaches}
The resulting segmentations from each of our experiments are compared to those
produced from two alternative segmentation approaches. The {\it single-atlas}
approach uses one atlas to segment a unlabelled subject by directly propagating
labels from the atlas by way of nonlinear registration.  We computed a
single-atlas segmentation for each image in the IBSR dataset from each of the
other 17 labelled images.  Similarly, we computed a segmentation for each image
using the {\it multi-atlas} approach, described above, using the other 17
images as the atlas library.  Additionally, we also varied the number of atlas
images used in the label fusion step by employing cross-correlation weighted
voting.  In total we evaluated approximately $52,000$ segmentations for the
work presented in this manuscript.

@@

\begin{figure}[h]
\begin{minipage}[b]{1.0\linewidth}
  \centering
  \includegraphics[width=\textwidth]{montage.png}
\end{minipage}
\caption{{\em Sample MAGeT brain segmentations.} 
Segmentations of a single subject are shown when three atlases are used, with
varying template library size $n$. Blue colouring represents agreement between
the gold-standard and MAGeT brain. Green colouring indicates false positive
voxels labelled by MAGeT brain, and red colouring indicates false negative
voxels.
} 
\label{montage}
\end{figure}

MAGeT brain shows distinct improvement in segmentation accuracy with increasing
atlas library size (Fig. \ref{results}), reaching its highest accuracy using 8
atlases (mean $\kappa = 0.775$).  This level of accuracy is within 2.0\% of the
multi-atlas approach (mean $\kappa = 0.791$) using 17 atlases. With only 3
atlases, MAGeT brain reaches to within 6.5\% accuracy of the multi-atlas
approach, and performs significantly better than the average single-atlas
performance. Surprisingly, our results do not show any significant improvements
in segmentation accuracy when applying cross-correlation weighted voting to
reduce the number of template labels being fused.

\begin{figure}
\begin{minipage}[b]{1.0\linewidth}
  \centering
  %\includegraphics[width=\textwidth]{results}
<<fig=TRUE, cached=TRUE, echo=FALSE, warnings=FALSE, >>=
library(ggplot2)
results = read.csv(file='data/results.csv')
xcorr_vote= subset(results, method=='xcorr' | method == 'multiatlas_xcorr_vote' )
qplot(x=top_n, y=k_hippocampus, data=xcorr_vote, colour=as.factor(num_atlases), geom=c('point'), alpha=I(0)) +
  geom_smooth(method="loess") + #line=1.5, formula=y~poly(x,2), method='lm') + 
  geom_hline(aes(yintercept = mean(subset(results, method=='naive')$k_hippocampus))) + 
  theme(legend.position="none") + 
  ylab('Mean similarity score (Kappa)') + 
  xlab('Number of template labels fused') +
  annotate("text", 14.3, 0.712, label="Single-atlas", size=4, hjust=0)+
  annotate("text", 15.3, 0.754, label="3 atlases", size=4, hjust=0)+
  annotate("text", 14.4, 0.766, label="4 atlases", size=4, hjust=0)+
  annotate("text", 13.5, 0.770, label="5 atlases", size=4, hjust=0)+
  annotate("text", 12.5, 0.773, label="6 atlases", size=4, hjust=0)+
  annotate("text", 11.5, 0.776, label="7 atlases", size=4, hjust=0)+
  annotate("text", 10.3, 0.779, label="8 atlases", size=4, hjust=0)+
  annotate("text", 14.5, 0.793, label="Multi-atlas", size=4, hjust=0)+
  annotate("text", 14.6, 0.787, label="17 atlases", size=4, hjust=0)+
  scale_colour_manual(values=c(
    "#D73027","#FC8D59","#FEE08B","#D9EF8B", "#91CF60","#1A9850", "#000000")) + 
  coord_cartesian(ylim=c(0.703,0.803), xlim=c(3,18)) + 
  scale_y_continuous(breaks=seq(0,1,0.01))

@  
\end{minipage}
\caption{
Mean performance of MAGeT brain with varying atlas library size and number of
template labels fused. Also shown is the mean multi-atlas performance when
using an atlas library of 17 images and varying the number of labels fused, as
well as the mean performance of single-atlas segmentations.  Data is fit with
LOESS local regression smoothing lines, and include a standard deviation
ribbon.
}
\label{results}
\end{figure}

@@

\subsection{Acknowledgements}
Computations were performed on the gpc supercomputer at the SciNet HPC
Consortium\cite{Loken2010}. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of Ontario;
Ontario Research Fund - Research Excellence; and the University of Toronto.

>>>>

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. Unfortunately, these
approaches often rely on a large number of such manually segmentated atlases
that take significant time and expertise to produce. We present an algorithm
for the automatic segmentation of the hippocampus that minimizes the number of
atlases needed while still achieving similar accuracy to multi-atlas
approaches.  We perform repeated random subsampling validation on the IBSR
dataset to compare our approach to basic multi-atlas segmentation using
the full IBSR dataset, and to single-atlas (model-based) segmentation. Our
results show that with only 8 input atlas, MAGeT brain can achieve to within
2.0\% segmentation accuracy of the basic multi-atlas approach using 17
input atlases (mean $\kappa = 0.775$ vs.  $\kappa = 0.791$).  These results
compare favourably to others who have used the IBSR data for validation. 
\end{abstract}

@@

Currently, the gold standard for neuroanatomical segmentation is manual
labelling by an expert human rater.  This is problematic for segmentation of
the hippocampus for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \cite{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true delineation of hippocampal anatomy in MR images is
disputed\cite{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol\cite{Jack2011}.

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders\cite{Mouiha2011}.  Additionally, it may be necessary
to use several different hippocampal definitions or, in fact, make
research-specific modifications in order to test hypotheses (for example,
\cite{Poppenk2011} found that overall hippocampal volume difference did not
predict recollection memory performance but by dividing the hippocampus into
anterior and posterior regions, a predictive volume difference was found).
Thus, while manual segmentation of the hippocampus is an important technique,
it may be a bottleneck for researchers or clinicians who do not have access to
the needed human expertise.

@@

The essential insight of generating a template library is not new.  Heckemann
\cite{Heckemann2006} compared generating a template library from a single atlas
to standard multi-atlas segmentation and found poor performance and so deemed
the approach as inviable.  The LEAP algorithm \cite{Wolz2010} proceeds by
iteratively segmenting the unlabelled image most similar to the atlas library
images and then incorporating the now-labelled image into the atlas library,
but requires 30 starting atlases.  The novelty of our method is to demonstrate
the possibility of producing comparable segmentation accuracy to these and
other multi-atlas-based methods while using significantly fewer manually
created atlases.

In previous work from our group \cite{Chakravarty2012}, we applied MAGeT brain
to the segmentation the striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. In the present work we extend our approach to the
hippocampus, a much more difficult structure to segment, and perform a thorough
validation over a range of atlas and template library sizes. Due to the small
number of atlases required, our method could easily accommodate different
hippocampal definitions. Our aim is not to improve on segmentation accuracy
beyond existing methods, but instead to provide a method that trades off manual
segmentation expertise for computational processing time while providing
sufficient accuracy for clinical and research applications.

@@

The segmentation approach we propose is best understood as an extension of
basic multi-atlas segmentation.  In multi-atlas segmentation, an atlas
library and unlabelled MR images are given as input.  Every atlas image is
nonlinearly registered to each unlabelled image, and then each atlas' labels
are propagated via the resulting transformations.  These labels are then fused
to produce a single, definitive segmentation by some label fusion method (e.g.
voxel-wise majority vote).

Our extension adds a preliminary stage in which a template library is
constructed from input images, and used in place of an atlas library in the
standard multi-atlas-based method.  To create the template library, labels from
each atlas image are propagated to each template library image via the
transformation resulting from a non-linear registration between pair of images.
As a result, each template library image has a label from each atlas.
Basic multi-atlas segmentation is then used to produce segmentations for the
entire set of unlabelled images (including those images used in the template
library).

@@

\subsubsection{Comparison Approaches}
The resulting segmentations from each of our experiments are compared to those
produced from two alternative segmentation approaches. The {\it single-atlas}
approach uses one atlas to segment a unlabelled subject by directly propagating
labels from the atlas by way of nonlinear registration.  We computed a
single-atlas segmentation for each image in the IBSR dataset from each of the
other 17 labelled images.  Similarly, we computed a segmentation for each image
using the {\it basic multi-atlas} approach, described above, using the
other 17 images as the atlas library.  Additionally, we also varied the number
of atlas images used in the label fusion step by employing cross-correlation
weighted voting.  In total we evaluated approximately $52,000$ segmentations
for the work presented in this manuscript.

@@

\begin{figure}[h]
\begin{minipage}[b]{1.0\linewidth}
  \centering
  \includegraphics[width=\textwidth]{montage.png}
\end{minipage}
\caption{{\small
Segmentations of a single subject are shown when three atlases are used, with
varying template library size $n$. Blue colouring represents agreement between
the gold-standard and MAGeT brain. Green colouring indicates false positive
voxels labelled by MAGeT brain (i.e. labelled voxels not appearing in the
gold-standard labels), and red colouring indicates false negative voxels (i.e.
voxels labelled in the gold-standard but not by MAGeT brain).
}} 
\label{montage}
\end{figure}

MAGeT brain achieves a level of segmentation accuracy that is to within 2.0\%
of the accuracy of the basic multi-atlas approach (Fig. \ref{results};
mean $\kappa = 0.775$ vs mean $\kappa = 0.791$) in the best case. Importantly,
to do this MAGeT brain only requires 8 manual segmented images whereas the
basic multi-atlas requires 17 atlases.  This represents a significant
savings in manual effort, and supports our contention that it is possible with
MAGeT brain to trade a small decrease in accuracy for a significant decrease in
the number of manual segmentations needed.

Surprisingly, our analysis does not show any significant improvements in
segmentation accuracy for either method when applying cross-correlation
weighted voting to reduce the number of template labels being fused.

\begin{figure}[t]
%\begin{minipage}[b]{1.0\linewidth}
  \centering
  %\includegraphics[width=\textwidth]{results}
<<fig=TRUE, cached=TRUE, echo=FALSE, warnings=FALSE >>=
library(ggplot2)
results = read.csv(file='data/results.csv')
xcorr_vote= subset(results, method=='xcorr' | method == 'multiatlas_xcorr_vote' )
qplot(x=top_n, y=k_hippocampus, data=xcorr_vote, colour=as.factor(num_atlases), geom=c('point'), alpha=I(0)) +
  geom_smooth(method="loess") + #line=1.5, formula=y~poly(x,2), method='lm') + 
  geom_hline(aes(yintercept = mean(subset(results, method=='naive')$k_hippocampus))) + 
  theme(legend.position="none") + 
  ylab('Mean similarity score (Kappa)') + 
  xlab('Number of template labels fused') +
  annotate("text", 14.3, 0.712, label="Single-atlas", size=4, hjust=0)+
  annotate("text", 15.3, 0.754, label="3 atlases", size=4, hjust=0)+
  annotate("text", 14.4, 0.766, label="4 atlases", size=4, hjust=0)+
  annotate("text", 13.5, 0.770, label="5 atlases", size=4, hjust=0)+
  annotate("text", 12.5, 0.773, label="6 atlases", size=4, hjust=0)+
  annotate("text", 11.5, 0.776, label="7 atlases", size=4, hjust=0)+
  annotate("text", 10.3, 0.779, label="8 atlases", size=4, hjust=0)+
  annotate("text", 13.5, 0.793, label="Basic multi-atlas", size=4, hjust=0)+
  annotate("text", 14.6, 0.787, label="17 atlases", size=4, hjust=0)+
  scale_colour_manual(values=c(
    "#D73027","#FC8D59","#FEE08B","#D9EF8B", "#91CF60","#1A9850", "#000000")) + 
  coord_cartesian(ylim=c(0.703,0.803), xlim=c(3,18)) + 
  scale_y_continuous(breaks=seq(0,1,0.01))

@  
%\end{minipage}
\caption{{\small
Mean performance of MAGeT brain with varying atlas library size and number of
template labels fused. Also shown is the mean basic multi-atlas
performance when using an atlas library of 17 images and varying the number of
labels fused, as well as the mean performance of single-atlas segmentations.
Data is fit with LOESS local regression smoothing. One standard deviation is
shown in grey.
}}
\label{results}
\end{figure}

@@

\section{Acknowledgements}
Computations were performed on the gpc supercomputer at the SciNet HPC
Consortium\cite{Loken2010}. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of Ontario;
Ontario Research Fund - Research Excellence; and the University of Toronto.

### 2012-11-01 paper.tex
Miriam and Julie comments incorporated. Changed all discussion of multi-atlas to "basic multi-atlas" to disambiguate. Squashed into four pages.

\begin{figure}[h]
\begin{minipage}[b]{1.0\linewidth}
  \centering
  \includegraphics[width=\textwidth]{montage.png}
\end{minipage}
\caption{{\em Sample MAGeT brain segmentations.} 
Segmentations of a single subject are shown when three atlases are used, with
varying template library size $n$. Blue colouring represents agreement between
the gold-standard and MAGeT brain. Green colouring indicates false positive
voxels labelled by MAGeT brain, and red colouring indicates false negative
voxels.
} 
\label{montage}
\end{figure}

MAGeT brain shows distinct improvement in segmentation accuracy with increasing
atlas library size (Fig. \ref{results}), reaching its highest accuracy using 8
atlases (mean $\kappa = 0.775$).  This level of accuracy is within 2.0\% of the
multi-atlas approach (mean $\kappa = 0.791$) using 17 atlases. With only 3
atlases, MAGeT brain reaches to within 6.5\% accuracy of the multi-atlas
approach, and performs significantly better than the average single-atlas
performance. Surprisingly, our results do not show any significant improvements
in segmentation accuracy when applying cross-correlation weighted voting to
reduce the number of template labels being fused.

\begin{figure}
\begin{minipage}[b]{1.0\linewidth}
  \centering
  %\includegraphics[width=\textwidth]{results}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\includegraphics[width=\maxwidth]{figure/unnamed-chunk-1} 
\end{knitrout}

\end{minipage}
\caption{
Mean performance of MAGeT brain with varying atlas library size and number of
template labels fused. Also shown is the mean multi-atlas performance when
using an atlas library of 17 images and varying the number of labels fused, as
well as the mean performance of single-atlas segmentations.  Data is fit with
LOESS local regression smoothing lines, and include a standard deviation
ribbon.
}
\label{results}
\end{figure}

>>>>

\begin{figure}[h]
\begin{minipage}[b]{1.0\linewidth}
  \centering
  \includegraphics[width=\textwidth]{montage.png}
\end{minipage}
\caption{{\small
Segmentations of a single subject are shown when three atlases are used, with
varying template library size $n$. Blue colouring represents agreement between
the gold-standard and MAGeT brain. Green colouring indicates false positive
voxels labelled by MAGeT brain (i.e. labelled voxels not appearing in the
gold-standard labels), and red colouring indicates false negative voxels (i.e.
voxels labelled in the gold-standard but not by MAGeT brain).
}} 
\label{montage}
\end{figure}

MAGeT brain achieves a level of segmentation accuracy that is to within 2.0\%
of the accuracy of the basic multi-atlas approach (Fig. \ref{results};
mean $\kappa = 0.775$ vs mean $\kappa = 0.791$) in the best case. Importantly,
to do this MAGeT brain only requires 8 manual segmented images whereas the
basic multi-atlas requires 17 atlases.  This represents a significant
savings in manual effort, and supports our contention that it is possible with
MAGeT brain to trade a small decrease in accuracy for a significant decrease in
the number of manual segmentations needed.

Surprisingly, our analysis does not show any significant improvements in
segmentation accuracy for either method when applying cross-correlation
weighted voting to reduce the number of template labels being fused.

\begin{figure}[t]
%\begin{minipage}[b]{1.0\linewidth}
  \centering
  %\includegraphics[width=\textwidth]{results}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\includegraphics[width=\maxwidth]{figure/unnamed-chunk-1} 
\end{knitrout}

%\end{minipage}
\caption{{\small
Mean performance of MAGeT brain with varying atlas library size and number of
template labels fused. Also shown is the mean basic multi-atlas
performance when using an atlas library of 17 images and varying the number of
labels fused, as well as the mean performance of single-atlas segmentations.
Data is fit with LOESS local regression smoothing. One standard deviation is
shown in grey.
}}
\label{results}
\end{figure}
